---
title: "Diatom-based ecological multivariate analyses"
author: "Vélez-Agudelo Camilo"
date: "8/3/2022"
output:
  html_notebook:
    code_folding: "show"
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(vegan)
library(rioja)
library(ggplot2)
library(ggpubr)
library(kableExtra)
library(gridExtra)
library(rstatix)
```

# **Diatom from Colorado River** #

## Loading Biological Data ##

This dataset contains diatom data of 15 modern surface sediments samples collected in the middle and lower basin of the Colora River, Patagonia.



```{r}
speciesRC <- readRDS("speciesRC.rds") ##Importing Diatom Data##
```

## Diatom diagram of common species ##

Remove diatom taxa where relative abundance is less than 4%

```{r}

## This code estimates the relative abundance of each diatom specie##
speciesRC.rel <- (decostand(
    speciesRC,
    "total",
    MARGIN = 1)
    ) * 100

speciesRC.rel2 <- colSums(speciesRC.rel) ##Calculate column sums

##Subset data keeping only diatom species with relative abundance greater than 4%##
speciesRC.red <- speciesRC.rel[, which(speciesRC.rel2 > 4)]

```

Cluster analyses of untransformed diatom data.

```{r Cluster.Analysis.RC, fig.show = "hold"}

dis.brayRC <- vegdist(
    x = speciesRC.rel,
    method = "bray"
    )

clustRC <- chclust(
    dis.brayRC, 
    method = "coniss"
    )

plot(clustRC,labels = NULL)

bstick(clustRC, plot = TRUE)

```

```{r Diatom.Diagram, fig.width=12}

color.d <- c(
    rep("darkseagreen4", 3), 
    rep("gold2", 10), 
    rep("chocolate3", 2)
    )

y.scale <- 1:15

p1 <- strat.plot(
    speciesRC.red,
    xLeft = 0.07,
    xRight = 1,
    yBottom = 0.05,
    yTop = 0.7,
    y.rev = T, 
    yvar = y.scale, 
    y.tks = y.scale, 
    scale.percent = T,
    plot.line = F,
    plot.bar = T,
    lwd.bar = 16,
    ylabel = "Sampling site",
    srt.xlabel = 45,
    xSpace = 0.005,
    wa.order = "bottomleft",
    x.pc.inc = 5,
    x.pc.lab = T,
    cex.ylabel = 1.5,
    cex.xlabel = 1,
    cex.yaxis = 0.8, 
    col.bar = color.d,
    col.line = color.d,
    clust = clustRC,
    tck = 5,
    tcl = 0.1,
    sep.bar = T,
    add.smooth = F,
    smooth.span = 0.1
    )

addClustZone(p1, clustRC, 4, col = "red")
```


## Diatom from Negro River ##

## Diatom from Chubut River ##
